---
title: "m565A4"
author: "Vinay Vaida"
date: "2023-10-19"
output:
  word_document: default
  pdf_document: default
---

```{r}
options(repos = c(CRAN = "https://cran.r-project.org/"))
#chooseCRANmirror(ind = 1)  # Choose the first mirror
```



1. (20) In this exercise, we will generate simulated data, and will then use
this data to perform best subset selection.

(a) Use the rnorm() function to generate a predictor $X$ of length
$n = 100$, as well as a noise vector $\epsilon$ of length $n = 100$.


```{r}
# Set a seed for reproducibility (for easy debugging of the code and long-term use)
set.seed(26)
# Generate a predictor variable X of length n = 100
X <- rnorm(n = 100, mean = 0, sd = 1)
head(X)
# Generate a noise vector epsilon of length n = 100
e <- rnorm(n = 100, mean = 0, sd = 1)
head(e)
```


(b) Generate a response vector $Y$ of length $n = 100$ according to
the model $Y = \beta_0 + \beta_1 X + \beta_2X^2 + \beta_3X^3 + \epsilon$,
where $\beta_0, \beta_1, \beta_2$, and $\beta_3$ are constants of your choice.

```{r}
# Define your chosen constants
beta_0 <- 3 # beta_0 value
beta_1 <- 6 #beta_1 value
beta_2 <- 9 #beta_2 value
beta_3 <- 12  #beta_3 value


# Calculate the response vector Y using the specified model
Y <- beta_0 + beta_1 * X + beta_2 * X^2 + beta_3 * X^3 + e

#First few values of Y
head(Y)
```


(c) Use the regsubsets() function to perform best subset selection
in order to choose the best model containing the predictors
$X,X^2,\ldots,X^{10}$. 

Note that you will need to use the data.frame() function to
create a single data set containing both $X$ and all the powers in consideration, and $Y$. Example: for $X^3$ use, in the data.frame, $X3=X^3$.

Use regsubsets() with nbest = 3,really.big = T, nvmax = 10

```{r}
# Create a data frame with X, X^2, ..., X^10 and Y
#mydata <- data.frame(X, X2 = X^2, X3 = X^3, X4 = X^4, X5 = X^5, X6 = X^6, X7 = X^7, X8 = X^8, X9 = X^9, X10 = X^10, Y)


X_df <- data.frame(
  X = X,
  X2 = X^2,
  X3 = X^3,
  X4 = X^4,
  X5 = X^5,
  X6 = X^6,
  X7 = X^7,
  X8 = X^8,
  X9 = X^9,
  X10 = X^10
)

Y_df <- data.frame(Y)

mydata<-cbind(Y_df, X_df)

# Load the 'leaps' library
install.packages("leaps")
library(leaps)

# Use regsubsets() to perform best subset selection
best <- regsubsets(Y ~ ., data = mydata, nbest = 3, really.big = TRUE, nvmax = 10)

head(mydata)
```

Before proceeding, follow structure in file modelselection Example: Heights. The goal is to find the best model obtained according to Cp, BIC, and adjusted $R^2$.

Report the coefficients of the best model obtained. 

See the results:
```{r}
# Get the summary of the best models
sbest <- summary(best)
sbest
```

See what is inside sbest:
```{r}
names(sbest)
```


To see the models and their Cp: (you can do the same with adjr2 and bic)
```{r}
cbind(sbest$which ,sbest$cp)
```

Follow the notes in modelselection Example: Heights to see how to select the best model according to the various metrics.
```{r}
mybestmodel<-function(Xnames,Yname,dataset,p,crit="bic"){
  if(crit=="Cp"){
    n<-dim(dataset)[1]
    fullMSE=summary(lm(as.formula(paste(Yname,"~.")),data=dataset))$sigma^2
  }
  varsel<-lapply(0:p, function(x) combn(p,x)) 
  
  modcrit<-numeric(p); form<-character(p)
  for(k in 1:p){
    s<-dim(varsel[[k+1]])[2] 
    tempform<-character(s); tempcrit<-numeric(s)
    for(j in 1:s){
      temp <- Xnames[varsel[[k+1]][,j]]
      tempform[j]<- ifelse(length(temp)>1,
      paste(temp, collapse = " + "), temp)
      tempform[j] <- paste(Yname, tempform[j],sep='~')
      tempmod<-lm(as.formula(tempform[j]),data=dataset)
      if(crit=="aic"){
        tempcrit[j] <- AIC(tempmod)
      }
      if(crit=="bic"){
        tempcrit[j] <- BIC(tempmod)
      }
      if(crit=="r2"){
        tempcrit[j] <- summary(tempmod)$adj
      }
      if(crit=="Cp"){
        tempcrit[j]<-sum(tempmod$resˆ2)/fullMSE+2*(k+1)-n
      }
    }
    # best model of size k
    if(crit %in% c("aic", "bic")){
      best<-which.min(tempcrit)
    }
    if(crit == "r2"){
      best<-which.max(tempcrit)
    }
    if(crit=="Cp"){
      best<-which.min(abs(tempcrit[j]-(k+1)))
    }
    form[k]<-tempform[best]
    modcrit[k]<-tempcrit[best]
  }
  if(crit %in% c("aic", "bic")){
    out<-form[which.min(modcrit)]
  }
  if(crit == "r2"){
    out<-form[which.max(modcrit)]
  }
  if(crit=="Cp"){
    out<-form[which.min(abs(modcrit[-p]-(2:p)))]
  }
  return(out)
}

p<-length(names(mydata))-1 
Xnames<-names(mydata)[-1]
Yname<-"Y"
dataset<-mydata
bicform<-mybestmodel(Xnames, Yname, mydata, p, crit="bic")
bicform

Criteria<-function(model){
  out<-data.frame(`p+1`=length(model$coef),
                  R2adj=summary(model)$adj,
                  AIC=AIC(model),
                  BIC=BIC(model))
  return(out)
}

modbic<-lm(as.formula(bicform), data=mydata)
modbic

aicform<-mybestmodel(Xnames, Yname, mydata, p, crit="aic")
aicform

modaic<-lm(as.formula(aicform), data=mydata)
modaic

cpform<-mybestmodel(Xnames, Yname, mydata, p, crit="Cp")
cpform

modCp<-lm(as.formula(cpform), data=mydata)
modCp

r2form<-mybestmodel(Xnames, Yname, mydata, p, crit="r2")
r2form

modr2<-lm(as.formula(r2form), data=mydata)
modr2
```

(d) Now find the best model using backward stepwise selection. 

```{r}
#backwards
modback<-step(lm(Y~.,data=mydata),trace=0, direction = "backward")
```


(e) Now fit a lasso model to the simulated data, again using $X,X^2,
\ldots, X^{10}$ as predictors. Use cross-validation to select the optimal
value of $\lambda$. Create plots of the cross-validation error as a function
of $\lambda$. Report the resulting coefficient estimates, and discuss the
results obtained.

```{r}
install.packages("glmnet")
library(glmnet)

X_matrix <- as.matrix(X_df)
Y_vector <- as.vector(Y_df$Y)

lasso.mod <- cv.glmnet(X_matrix, Y_vector, alpha = 1)
plot(lasso.mod)

best_lambda <- lasso.mod$lambda.min
cat("Optimal lambda:", best_lambda, "\n")

lasso_coef <- coef(lasso.mod, s = best_lambda)
cat("Coefficient estimates for the optimal lambda:\n")
print(lasso_coef)
```



(f) Display the adjR^2, BIC and AIC for all the models

```{r}
rbind(bicm=Criteria(modbic),
      aicm=Criteria(modaic),
      adjr2m=Criteria(modr2),
      cpm=Criteria(modCp),
      bsel=Criteria(modback)
)
```


(g) Now use cross validation to select the final model.

```{r}
cv.lm <- function(data, formulae, nfolds = 5) {
  data <- na.omit(data) # remove missing values
  formulae <- sapply(formulae, as.formula)
  n <- nrow(data)
  fold.labels <- sample(rep(1:nfolds, length.out = n))
  mses <- matrix(NA, nrow = nfolds, ncol = length(formulae))
  colnames <- as.character(formulae)
  for (fold in 1:nfolds) {
    test.rows <- which(fold.labels == fold)
    train <- data[-test.rows, ]
    test <- data[test.rows, ]
    for (form in 1:length(formulae)) {
      current.model <- lm(formula = formulae[[form]], data = train)
      predictions <- predict(current.model, newdata = test)
      test.responses <- eval(formulae[[form]][[2]], envir = test)
      test.errors <- test.responses - predictions
      mses[fold, form] <- mean(test.errors^2)
    }
  }
  return(colMeans(mses))
}

set.seed(1999)
formulae<-c(formula(modbic),
            formula(modaic),
            formula(modr2),
            formula(modCp),
            formula(modback))

mse<-cv.lm(data=mydata, formulae, nfolds = 5)
print(mse)
```
From the above obtained values, we observe that the model with AIC metric has the least error value. Therefore, the model with AIC is the final model.

Following are the Error values in order:

modelAIC > modelBIC > modback > modr2 > modCp

2. (30) The perils of post-selection inference, and data splitting to the rescue.

(a) Generate a 1000 x 101 array, where all the entries are IID standard 
Gaussian variables. 

We'll call the first column the response variable $Y$, 
and the others the predictors $X_1,\ldots,X_{100}$. 

```{r}
mydata<-as.data.frame(matrix(rnorm(1000*101),ncol=101))
names(mydata)<-c("Y",paste("X",1:100,sep=""))
mydata[1:3,1:5]
```


By design, there is no true relationship between the response and the predictors (but all the usual linear-Gaussian-modeling assumptions hold).

(b) 
  (i) Estimate the model $Y=\beta_0+\beta_1X_1+\beta_{50}X_{50}+\epsilon$
    Extract the p-value for the **F test** of the whole model. 
```{r}
getpvalue <- function(model, mydata) {
  model <- lm(model, data = mydata)
  sum<-summary(model)
  pvalue<-pf(sum$fstatistic[1], sum$fstatistic[2], sum$fstatistic[3], lower.tail=FALSE)
  return (pvalue)
}

getpvalue(Y~X1+X50, mydata)
```


  (ii) Repeat the simulation (steps a and b), estimation and testing 
    100 times, and plot the histogram of the p-values. 
    What does it look like? What should it look like?
```{r}
 pvalue_list <- numeric(100)
for (i in 1:100){
  gendata<-as.data.frame(matrix(rnorm(1000*101),ncol=101))
  names(gendata)<-c("Y",paste("X",1:100,sep=""))
  pvalue_list[i]<-getpvalue(Y~X1+X50, gendata)
}

# Create a histogram
 hist(pvalue_list, main = "Histogram", xlab = "P-value", ylab = "Frequency", col = "pink", border = "black")
```



(c) 
  (i) From step (a), use the step function to select a linear model by        backward stepwise selection. 
```{r}
final_model <- step(lm(Y~., mydata), direction = "backward", trace = 0)
model_backward <- function(model, data){
  final_model <- step(model, direction = "backward", trace = 0)
  sum<-summary(final_model)
  pvalue<-pf(sum$fstatistic[1], sum$fstatistic[2], sum$fstatistic[3], lower.tail=FALSE)
  return (pvalue)

}
model <- lm(Y~., data = mydata)
model_backward(model, mydata)
```
  

  (ii) Extract the p-value for the **F-test** of the selected model. 
    Repeat 100 times and plot the histogram of p-values. 
    Explain what's going on.
```{r}
pvalue_list_back <- list()
for (i in 1:100){
  gendata<-as.data.frame(matrix(rnorm(1000*101),ncol=101))
  names(gendata)<-c("Y",paste("X",1:100,sep=""))
  model <- lm(Y~., data = gendata)
  pvalue_list_back[i]<-model_backward(model, gendata)
}
combined_data <- unlist(pvalue_list_back)
# Create a histogram
hist(combined_data, main = "Histogram", xlab = "P-value", ylab = "Frequency", breaks = 100, col = "pink", border = "black")

```




(d) Again, 

  (i) use step() to select a model based on one random 1000x101 array.        This acts as a training dataset. Compute the in-sample error.
  
```{r}
traindata<-as.data.frame(matrix(rnorm(1000*101),ncol=101))
names(traindata)<-c("Y",paste("X",1:100,sep=""))

trainmodel <- lm(Y~., data = traindata)
step_model <- step(trainmodel, trace = 0)

summary(step_model)
```
```{r}
in_sample_error <- mean(residuals(step_model)^2)

cat("In-sample Mean Squared Error:", in_sample_error, "\n")
```
  

  (ii) Now re-estimate the selected model on a new 1000 x 101 array, 
    and extract the new p-value. This acts as a testing dataset. 
    Compute the generalization (prediction) error.
    
```{r}
testdata<-as.data.frame(matrix(rnorm(1000*101),ncol=101))
names(testdata)<-c("Y",paste("X",1:100,sep=""))

sum<-summary(step_model)
pvalue<-pf(sum$fstatistic[1], sum$fstatistic[2], sum$fstatistic[3], lower.tail=FALSE)

pvalue

predictions_test <- predict(step_model, newdata = testdata)

generalization_error <- mean((testdata$Y - predictions_test)^2)

cat("Generalization Error:", generalization_error, "\n")
```
    

(iii) Repeat 100 times, with new selection and inference sets each time,
  and plot the histogram of p-values, the histograms of in-sample errors
  and the histogram of generalization errors.
```{r}
isr_list <- numeric(100)
gen_list <- numeric(100)
pvalue_list_step <- numeric(100)

for(i in 1:100){
  train_gen_data<-as.data.frame(matrix(rnorm(1000*101),ncol=101))
  names(train_gen_data)<-c("Y",paste("X",1:100,sep=""))
  
  test_gen_data<-as.data.frame(matrix(rnorm(1000*101),ncol=101))
  names(test_gen_data)<-c("Y",paste("X",1:100,sep=""))
  
  genmodel <- lm(Y~., data = train_gen_data)
  final_gen_model <- step(genmodel, trace = 0)
  
  sum<-summary(final_gen_model)
  
  predictions_test <- predict(final_gen_model, newdata = test_gen_data)
  
  pvalue_list_step[i]<-pf(sum$fstatistic[1], sum$fstatistic[2], sum$fstatistic[3], lower.tail=FALSE)
  
  isr_list[i] <- mean(residuals(final_gen_model)^2)
  
  gen_list[i] <- mean((test_gen_data$Y - predictions_test)^2)
  
}

hist(pvalue_list_step, main = "Histogram", xlab = "P-value", ylab = "Frequency", col = "lightblue", border = "black")

hist(isr_list, main = "Histogram", xlab = "in-sample-error", ylab = "Frequency", col = "lightblue", border = "black")

hist(gen_list, main = "Histogram", xlab = "generalization error", ylab = "Frequency", col = "lightblue", border = "black")

```



